Quantum H theorem, variational entropy, and relaxation processes

### 2020-11-06 notasparaarticulo.tex
some parts were modified. Those parts are in red

\textcolor{red}{In the literature, the theoretical basis that allows us to describe equilibrium systems in the classical scheme is well-established. The procedures to describe equilibrium systems are well-known and work in almost every system in nature (Thermodynamics for a phenomenological description and statistical mechanics for a microscopic description). In contrast, a unified mathematical model to describe the evolution of any system to an equilibrium state has not developed. Nevertheless, developments can be found in the literature and describe the evolution to equilibrium to almost any system (the internal behavior of glasses \cite{cristal1} and the entropy measurement \cite{cristal2, cristal3} are some examples where the recent developments do not work.) Examples of those developments can be the Onsager formulation (called linear thermodynamics) \cite{kei, onsager}, and the Kinetic Theory of dilute gases developed by Boltzmann.\\

@@

The first part of this work consists of obtaining the classical equilibrium distribution function (Maxwell-Boltzmann distribution), the Boltzmann transport equation from the Boltzmann $H$ functional (through variational procedures), and the proof of the Boltzmann's $H$ theorem (without using the Boltzmann transport equation). The second part consists of obtaining the quantum equilibrium distribution functions (Bose-Einstein and Fermi-Dirac distributions), the quantum analogous to the Boltzmann transport equation from a new functional named as \textit{variational entropy}, and the proof of the quantum analogous Boltzmann's $H$ theorem. The third part consists of analyzing the corresponding principle between the \textit{variational entropy} and Boltzmann's $H$ functional.}

@@

@@

\textcolor{red}{The Boltzmann's $H$ functional is defined as

@@

Inspired by the Hamilton principle, $H'$ will be maximized with those restrictions to obtain the equilibrium distribution function for a classical gas using the Lagrange multipliers method obtaining}

@@

where $L+P=K$. \textcolor{red}{The group of $L$ cells represents such cells that have an excess of particles or/and energy (from the mean value in equilibrium). In contrast, the group of $P$ cells represents such cells that have to miss particles or/and energy. Besides, $\dot{g}^{+}_{J}$ and $\dot{g}^{-}_{J}$  represent the change on the deviations on cells that have excess or/and missing of particles or energy respectively while $g^{+}_{J}$ and $g^{-}_{J}$ represent the deviation on cells that have excess or/and missing of particles or energy respectively.\\
On the one hand, $\dot{g}^{+}_{J}<0$ describes the loss of particles and/or energy and so, $g^{+}_{J}>0$. On the other hand, $\dot{g}^{-}_{J}>0$ describes the gain of particles and/or energy and therefore $g^{-}_{J}<0$. \\

@@

\textcolor{red}{We can observe that $\bar{f}(v)$ in (\ref{classicalH4}) is always positive. Also the deviation and its derivative are positive, then all the expression is positive. However, the global sign makes the derivative of the $H'$ with respect to time is always less to zero.} With this we proved that $\frac{dH'}{dt}<0$. If the system is in equilibrium, $g(r,v,t)=\dot g(r,v,t)=0$, therefore $\frac{dH'}{dt}=0$.\\
\textcolor{red}{Joining both results, we can say the following statement:\\

@@

This statement will be the classical $H$ theorem with inhomogeneities.}

@@

In the next subsection, we obtain the Boltzmann transport equation from the Boltzmann's $H$ functional.

@@

\subsection{The Boltzmann transport equation}
The Boltzmann $H$ functional is defined as
    H=\int_{}^{}f(\vec{v},t)\ln f(\vec{v},t)\mathrm{d}^{3}v,
where $f$ is the spatial-homogeneous distribution function of the system and $v$ is the velocity of the particles defined in the $\mu$-space. \textcolor{red}{In order to obtain the Boltzmann transport equation, is necessary to consider that the distribution function is not homogeneous. To obtain the spatial dependence in Boltzmann's transport equation, we need to assume that cells are too small, such that they can be considered as differential volume elements. That is
    f_M(\vec{v},t)    \ \ \ \Rightarrow \ \ \ f(\vec{r},\vec{v},t),
where $r$ is the position of the particles defined in the $\mu$-space.} Because of that, we use the following functional
    H_{mod}=\int f(\vec{r},\vec{v},t) \ln f(\vec{r},\vec{v},t) \mathrm{d}^{3}v,
that involves the position dependence.
%\begin{equation}
%    H'=\int_{}^{}\int_{}^{}f(\vec{r},\vec{v},t)\ln f(\vec{r},\vec{v},t)\mathrm{d}^{3}v \mathrm{d}^{3}r.
%\end{equation}
 %This dependence is necessary to obtain the complete Boltzmann transport equation.\\
The following relation must be held in equilibrium 
    \delta H_{mod}=0.\label{variation1}
This due to $H$ functional in equilibrium is proportional to entropy and a maximum.\\ 
Assuming that $\vec{r}=\vec{r}(t)$ and $\vec{v}=\vec{v}(t)$, (\ref{variation1}) casts into
    0=\left( 1+\ln f \right)\left( \frac{\partial f}{\partial t}+\dot{\vec{r}}\cdot \nabla f+\dot{\vec{v}}\cdot \nabla_{\vec{v}} f \right).
The previous relation holds if $(1+\ln f)=0$ or $(\frac{\partial f}{\partial t}+\dot{\vec{r}}\cdot \nabla f+\dot{\vec{v}}\cdot \nabla_{\vec{v}} f)=0$.\\
Assuming that $(1+\ln f)=0$, we find that $f$ is a constant and consequently, $(\frac{\partial f}{\partial t}+\dot{\vec{r}}\cdot \nabla f+\dot{\vec{v}}\cdot \nabla_{\vec{v}} f)=0$. On the other hand, if $f$ is not a constant, the equation of motion will be
    \frac{\partial f}{\partial t}+\dot{\vec{r}}\cdot \nabla f+\dot{\vec{v}}\cdot \nabla_{\vec{v}} f=0.\label{transport}
The equation (\ref{transport}) is hold only in equilibrium. If the system is out of equilibrium, the variation is not equal to zero. As a consequence of the classical $H$ theorem, the variation is less to zero. In conclusion, it is necessary to include a term in (\ref{transport}). This term, as in \cite{huang}, will be the collision term and will be expressed in such form as
    \frac{\partial f}{\partial t}+\vec{v}\cdot \nabla f+\frac{\vec{F}}{m}\cdot \nabla_{\vec{v}} f=\left( \frac{df}{dt} \right)_{coll}, \label{classictransport}
where $\vec{F}$ is an external force, and $m$ is the mass of the particles.\\
\textcolor{red}{The equation (\ref{classictransport}) corresponds to the Boltzmann transport equation. \\
We observed that the variational method and the inclusion of a non-spatial homogeneous distribution function are consistent with the results obtained from Statistical Mechanics. In other words, we proved the classical $H$ theorem and we obtained the Boltzmann transport equation from the Boltzmann $H$ functional.\\
In the next section, we will apply the procedure used in this section to a new functional that describes a quantum gas (bosons or fermions).}

@@

\section{Quantum scheme}
\subsection{Variational entropy, Bose-Einstein, and Fermi-Dirac distribution}

\textcolor{red}{We begin by defining the form of the functional 

@@

  \textcolor{red}{We obtained some results about equilibrium systems. Those results are consistent with the information obtained from Statistical Mechanics. Now we proceed to analyze the behavior of derivative of the variational entropy $\Ss$ with respect to time in order to find the quantum analogous to Boltzmann's $H$ theorem.}

@@

This distribution function \textcolor{red}{must satisfy the local equilibrium hypothesis in each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the non-homogeneous distribution function assumption in the total volume ($f_{nM}$ must be different among all cells)}, and the following restrictions

@@

  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. \textcolor{red}{The previous conditions and restrictions are the discrete analogous case of the classical (continuous) case. Also, this system will suffer, as in the classical case, a \textit{relaxation process}.\\
  The quantities $\Delta_M,\delta_M$ are sufficiently big to be different from fluctuations in the system but sufficiently small to consider the system not far from the equilibrium state, as in the classical case.}  \\

@@

where $L+P=K$. Besides, $\dot{g}^{+}_{nJ}$ and $\dot{g}^{-}_{nJ}$  represent the change on the deviations on cells that have excess or/and missing of particles or energy respectively while $g^{+}_{nJ}$ and $g^{-}_{nJ}$ represent the deviation on cells that have excess or/and missing of particles or energy respectively.\\
On the one hand, $\dot{g}^{+}_{nJ}<0$ describes the loss of particles and/or energy and so, $g^{+}_{nJ}>0$. On the other hand, $\dot{g}^{-}_{nJ}>0$ describes the gain of particles and/or energy and therefore $g^{-}_{nJ}<0$. \\

@@

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\Ss}{dt}=0$. Then, we can say that any ideal quantum gas perturbed always evolves to the equilibrium state, such as the classical $H$ theorem. \textcolor{red}{Then, we obtained the quantum version of the Boltzmann's $H$ theorem including inhomogeneities in the system.\\
Now, we proceed to obtain the quantum analogous of the Boltzmann transport equation from the variational entropy $\Ss$ through variational method.}

@@

  \subsection{The quantum analogous of the Boltzmann transport equation}
Consider that the distribution function $f_{nM}$ is a function of the local number of particles $\mathcal{N}_M$, the local number of particles in a quantum energy state $\mathcal{N}_{M}^{n}$ and the time $t$ as follow
\begin{equation}
    f=f(\mathcal{N}_M,\mathcal{N}_M^{n},t),
\end{equation}
where $\mathcal{N}_{M}^{n}$ is defined as
\begin{equation}
    \mathcal{N}_{M}^{n}= f_{nM} \delta \epsilon_n.
\end{equation}
Besides the local number of particles is a function of time $\mathcal{N}_J=\mathcal{N}_J(t)$ and the local number of particles in a quantum energy state is a function of time and the local number of particles $\mathcal{N}_J^{j}=\mathcal{N}_J^{j}(\mathcal{N}_M,t)$. This is due to when a cell transfer particles to another cell, the particles in some defined state energy levels must change.\\
In order to obtain the analogous transport equation to this system, we apply the variation method to the functional $\Ss$. Due to $\Ss$ in equilibrium is proportional to entropy, and therefore a maximum, the variation must be zero. Then the equation of motion is
\begin{equation}
    0=\sum_{M=1}^{K} \sum_n \ln \left[ \frac{f_{nM}}{1\mp f_{nM}} \right] \left[ \frac{\partial f_{nM}}{\partial t}+\frac{\partial f_{nM}}{\partial \mathcal{N}_J}\dot{\mathcal{N}}_J+\frac{\partial f_{nM}}{\partial \mathcal{N}_J^{j}}\left( \frac{\partial \mathcal{N}_J^{j}}{\partial t}+\frac{\partial \mathcal{N}_J^{j}}{\partial \mathcal{N}_L}\dot{\mathcal{N}}_L \right) \right]. \label{transportequation1}
\end{equation}
This equation is valid in equilibrium. When the system is out of equilibrium, the variation is different from zero. Due to this, an expression has to be included in (\ref{transportequation1}). This term, in analogy to the Boltzmann transport equation, will be the collision term and will be expressed as $\left( \frac{df_{nM}}{dt} \right)_{coll}$. Moreover, we propose that this term contributes to the equation of motion per each cell.\footnote{\textcolor{red}{The collision term is only in a defined cell and local group of energy. If we are describing the total system, we need to include all contributions (each cell and energy group).}} Therefore, (\ref{transportequation1}) casts into
\begin{eqnarray}
    \sum_{M=1}^{K}\sum_n \left( \frac{df_{nM}}{dt} \right)_{coll}&=&\sum_{M=1}^{K} \sum_n \ln \left[ \frac{f_{nM}}{1\mp f_{nM}} \right] \nonumber \\
    &&\times \left[ \frac{\partial f_{nM}}{\partial t}+\frac{\partial f_{nM}}{\partial \mathcal{N}_J}\dot{\mathcal{N}}_J+\frac{\partial f_{nM}}{\partial \mathcal{N}_J^{j}}\left( \frac{\partial \mathcal{N}_J^{j}}{\partial t}+\frac{\partial \mathcal{N}_J^{j}}{\partial \mathcal{N}_L}\dot{\mathcal{N}}_L \right) \right].  \nonumber \\ \label{transportequation2}
\end{eqnarray}
The equation (\ref{transportequation2}) is the time evolution equation to a quantum ideal gas.

@@

\section{Classical limit}
\textcolor{red}{%The variational entropy $H$ that was defined in (\ref{entropy2}) can be reduce to the Boltzmann's $H$ functional in the appropriated limit.\\
We begin defining the distribution function as the number of particles in a defined cell and an energy level $N_{nM}$ inside a volume $\de V_M \delta \epsilon_n$ as
\begin{equation}
    f_{nM}=\frac{N_{nM}}{ \de V_M \delta \epsilon_{n} }.
\end{equation}
Using this definition on the variational entropy (\ref{entropy2}) we obtain
\begin{eqnarray}
    H&=& \sum_{M=1}^{K} \sum_n
    \left[  
           \frac{N_{nM}}{ \de V_M\delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{nM}}{ \de V_M\delta \epsilon_{n}}
           \right)\pm 
           \left(  
                  1\mp \frac{N_{nM}}{ \de V_M \delta \epsilon_{n}}
           \right) \ln 
           \left(  
                   1\mp \frac{N_{nM}}{ \de V_M \delta \epsilon_{n}}
           \right)
    \right] \de V_M \delta \epsilon_{n}. \nonumber \\
    \label{h-quantic} 
\end{eqnarray}
In the classic limit, the particles occupy high energy levels, and the number of particles in most of the groups of state $n$ is small compared with the number of the elementary states $\epsilon_n$ of the group, that is
\begin{equation}
    \frac{N_{nM}}{ \de V_M \delta \epsilon_{n} } \approx 0,
\end{equation}
and with the previous approximation, the second term in (\ref{h-quantic}) is negligible, then (\ref{h-quantic}) yields
\begin{eqnarray}
    H&=& \sum_{M=1}^{K} \sum_n
    \left[  
           \frac{N_{nM}}{ \de V_M \delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{nM}}{ \de V_M \delta \epsilon_{n}}
           \right)
    \right] \de V_M \delta \epsilon_{n}. \label{h-quantic2}
\end{eqnarray}
In the case when the system is spatial-homogeneous, the definition of the distribution function is 
\begin{equation}
    f_{n}=\frac{N_{n}}{ \delta \epsilon_{n} },
\end{equation}
and using the same limit we obtain
\begin{eqnarray}
    H&=& \sum_{M=1}^{K} \sum_n
    \left[  
           \frac{N_{n}}{ \delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{n}}{ \delta \epsilon_{n}}
           \right)
    \right]  \de V_M \delta \epsilon_{n} = V \sum_n
    \left[  
           \frac{N_{n}}{ \delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{n}}{ \delta \epsilon_{n}}
           \right)
    \right] \delta \epsilon_{n} \nonumber \\
    &=& V \sum_n \left[N_n \ln N_n - N_n \ln \delta \epsilon_n  \right]=\de V_M K H_{Boltz},\label{h-quantic4}
\end{eqnarray}
where $B_{Boltz}$ now is (\ref{CH2}) expressed is the form of (\ref{reduce-h}) shown by Tolman.\footnote{See \cite{tolman} eq. (102.6) for more information.}\\
We can observe that if the number of particles is too big compared to the number of cells $K$, then we can conclude that the sum of the local variational entropy is equal to the variational entropy of the entire system.}

@@

\section{Conclusions}
\textcolor{red}{We conclude the following statements:\\
\begin{enumerate}
    \item Any system (classical or quantum) out of equilibrium (including inhomogeneities) evolves to the equilibrium state without using the transport equation as long as the local equilibrium hypothesis holds.
    \item We can obtain the transport equation from the $H$ functional.
    \item The variational entropy is proportional to the Boltzmann's $H$ functional in the limit of the high energy levels.
    \item We find an expression for the entropy to quantum gases.
    \end{enumerate}
  }
%  \begin{equation}
%      \mathcal{H}
%  \end{equation}

>>>>

In the literature, the theoretical basis that allows us to describe equilibrium systems in the classical scheme is well-established. The procedures to describe equilibrium systems are well-known and work in almost every system in nature (Thermodynamics for a phenomenological description and statistical mechanics for a microscopic description). In contrast, a unified mathematical model to describe the evolution of any system to an equilibrium state has not developed. Nevertheless, developments can be found in the literature and describe the evolution to equilibrium to almost any system (the internal behavior of glasses \cite{cristal1} and the entropy measurement \cite{cristal2, cristal3} are some examples where the recent developments do not work.) Examples of those developments can be the Onsager formulation (called linear thermodynamics) \cite{kei, onsager}, and the Kinetic Theory of dilute gases developed by Boltzmann.\\

@@

\textcolor{red}{The first section of this work consists of obtaining some classical results from Statistical mechanics (the equilibrium distribution function, the Boltzmann transport equation, and an alternative proof of the classical $H$ theorem). The second section consists of obtaining the same results from quantum mechanics. The third consists of analyzing the corresponding principle between the \textit{variational entropy} and Boltzmann's $H$ functional.}

@@

\textcolor{red}{This section will be divided in three subsections, the first will be dedicated to obtain the equilibrium distribution function (Maxwell-Boltzmann distribution) from the Boltzmann's $H$ functional applying variational procedure.}

@@

The Boltzmann's $H$ functional is defined as

@@

Inspired by the Hamilton principle, $H'$ will be maximized with those restrictions to obtain the equilibrium distribution function for a classical gas using the Lagrange multipliers method obtaining

@@

where $L+P=K$. \textcolor{red}{The group of $L$ cells is such that cells have an excess of particles or/and energy (from the mean value in equilibrium). In contrast, the group of $P$ cells is such that cells have to miss particles or/and energy (from the mean value in equilibrium). Besides, $\dot{g}^{+}_{J}$ represents the change on the deviation on cells that have an excess of particles or energy while $\dot{g}^{-}_{J}$  represents the change on the deviations on cells that have missing particles or energy. 
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents the value of the deviation on cells that have missing particles or energy.\\
Also, on the one hand, $\dot{g}^{+}_{J}<0$ describes the loss of particles and/or energy and so, $g^{+}_{J}>0$. On the other hand, $\dot{g}^{-}_{J}>0$ describes the gain of particles and/or energy and therefore $g^{-}_{J}<0$. \\

@@

We can observe that $\bar{f}(v)$ in (\ref{classicalH4}) is always positive. Also the deviation and its derivative are positive, then all the expression is positive. However, the global sign makes the derivative of the $H'$ with respect to time is always less to zero. With this we proved that $\frac{dH'}{dt}<0$. If the system is in equilibrium, $g(r,v,t)=\dot g(r,v,t)=0$, therefore $\frac{dH'}{dt}=0$.\\
Joining both results, we can say the following statement:\\

@@

This statement will be the classical $H$ theorem with inhomogeneities.

@@

%\textcolor{red}{The last result we want to obtain is the Boltzmann transport equation from the Boltzmann $H$ functional. This procedure is developed in the next subsection.}

@@

\subsection{The Boltzmann transport equation}
The Boltzmann $H$ functional is defined as
\begin{equation}
    H=\int_{}^{}f(\vec{v},t)\ln f(\vec{v},t)\mathrm{d}^{3}v,
\end{equation}
where $f$ is the spatial-homogeneous distribution function of the system and $v$ is the velocity of the particles defined in the $\mu$-space. In order to obtain the Boltzmann transport equation, is necessary to consider that the distribution function is not homogeneous. To obtain the spatial dependence in Boltzmann's transport equation, we need to assume that cells are too small, such that they can be considered as differential volume elements. That is
\begin{equation}
    f_M(\vec{v},t)    \ \ \ \Rightarrow \ \ \ f(\vec{r},\vec{v},t),
\end{equation}
where $r$ is the position of the particles defined in the $\mu$-space. Because of that, we use the following functional
\begin{equation}
    H_{mod}=\int f(\vec{r},\vec{v},t) \ln f(\vec{r},\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
that involves the position dependence.
%\begin{equation}
%    H'=\int_{}^{}\int_{}^{}f(\vec{r},\vec{v},t)\ln f(\vec{r},\vec{v},t)\mathrm{d}^{3}v \mathrm{d}^{3}r.
%\end{equation}
 %This dependence is necessary to obtain the complete Boltzmann transport equation.\\
\\
The following relation must be held in equilibrium 
\begin{equation}
    \delta H_{mod}=0.\label{variation1}
\end{equation}
This due to $H$ functional in equilibrium is proportional to entropy and a maximum.\\ 
Assuming that $\vec{r}=\vec{r}(t)$ and $\vec{v}=\vec{v}(t)$, (\ref{variation1}) casts into
\begin{equation}
    0=\left( 1+\ln f \right)\left( \frac{\partial f}{\partial t}+\dot{\vec{r}}\cdot \nabla f+\dot{\vec{v}}\cdot \nabla_{\vec{v}} f \right).
\end{equation}
The previous relation holds if $(1+\ln f)=0$ or $(\frac{\partial f}{\partial t}+\dot{\vec{r}}\cdot \nabla f+\dot{\vec{v}}\cdot \nabla_{\vec{v}} f)=0$.\\
Assuming that $(1+\ln f)=0$, we find that $f$ is a constant and consequently, $(\frac{\partial f}{\partial t}+\dot{\vec{r}}\cdot \nabla f+\dot{\vec{v}}\cdot \nabla_{\vec{v}} f)=0$. On the other hand, if $f$ is not a constant, the equation of motion will be
\begin{equation}
    \frac{\partial f}{\partial t}+\dot{\vec{r}}\cdot \nabla f+\dot{\vec{v}}\cdot \nabla_{\vec{v}} f=0.\label{transport}
\end{equation}
The equation (\ref{transport}) is hold only in equilibrium. If the system is out of equilibrium, the variation is not equal to zero. As a consequence of the classical $H$ theorem, the variation is less to zero. In conclusion, it is necessary to include a term in (\ref{transport}). This term, as in \cite{huang}, will be the collision term and will be expressed in such form as
\begin{equation}
    \frac{\partial f}{\partial t}+\vec{v}\cdot \nabla f+\frac{\vec{F}}{m}\cdot \nabla_{\vec{v}} f=\left( \frac{df}{dt} \right)_{coll}, \label{classictransport}
\end{equation}
where $\vec{F}$ is an external force, and $m$ is the mass of the particles.\\
\\
The equation (\ref{classictransport}) corresponds to the Boltzmann transport equation. \\
We observed that the variational method and the inclusion of a non-spatial homogeneous distribution function are consistent with the results obtained from Statistical Mechanics. In other words, we proved the classical $H$ theorem and we obtained the Boltzmann transport equation from the Boltzmann $H$ functional.\\
\textcolor{red}{In contrast, there is not quantum analogous of the Boltzmann transport equation. Even more, the analogous quantum $H$ theorem is not clear on the spatial-homogeneous distribution hypothesis. To obtain those results in the quantum mechanic scheme, the next section will develop the same procedure used in this section.}

@@

\section{Quantum scheme}
\textcolor{red}{This section will be divided into three subsections (as in the previous section). The first part consists of obtaining the quantum equilibrium distribution functions (Bose-Einstein and Fermi-Dirac distributions) from a new functional named \textit{variational entropy}. In the second part, we prove the quantum analogous of the $H$ theorem using the same way as in the previous section. Finally, in the third part, we obtain an equation of motion for the distribution function from the \textit{variational entropy}. This equation can be treated as the quantum analogous of the Boltzmann transport equation.} 
\subsection{Variational entropy, Bose-Einstein, and Fermi-Dirac distribution}

We begin by defining the form of the functional 

@@

 % \textcolor{red}{We obtained some results about equilibrium systems. Those results are consistent with the information obtained from Statistical Mechanics. Now we proceed to analyze the behavior of derivative of the variational entropy $\Ss$ with respect to time in order to find the quantum analogous to Boltzmann's $H$ theorem.}

@@

This distribution function must satisfy the local equilibrium hypothesis in each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the non-homogeneous distribution function assumption in the total volume ($f_{nM}$ must be different among all cells), and the following restrictions

@@

  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. The previous conditions and restrictions are the discrete analogous case of the classical (continuous) case. Also, this system will suffer, as in the classical case, a \textit{relaxation process}.\\
  The quantities $\Delta_M,\delta_M$ are sufficiently big to be different from fluctuations in the system but sufficiently small to consider the system not far from the equilibrium state, as in the classical case.  \\

@@

where $L+P=K$. \textcolor{red}{Besides, $\dot{g}^{+}_{J}$ represents the change on the deviation on cells that have an excess of particles or energy while $\dot{g}^{-}_{J}$  represents the change on the deviations on cells that have missing particles or energy. 
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents the value of the deviation on cells that have missing particles or energy.\\
Also, on the one hand}, $\dot{g}^{+}_{nJ}<0$ describes the loss of particles and/or energy and so, $g^{+}_{nJ}>0$. On the other hand, $\dot{g}^{-}_{nJ}>0$ describes the gain of particles and/or energy and therefore $g^{-}_{nJ}<0$. \\

@@

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\Ss}{dt}=0$. Then, we can say that any ideal quantum gas perturbed always evolves to the equilibrium state, such as the classical $H$ theorem. Then, we obtained the quantum version of the Boltzmann's $H$ theorem including inhomogeneities in the system.\\
%Now, we proceed to obtain the quantum analogous of the Boltzmann transport equation from the variational entropy $\Ss$ through variational method.

@@

  \subsection{The quantum analogous of the Boltzmann transport equation}
Consider that the distribution function $f_{nM}$ is a function of the local number of particles $\mathcal{N}_M$, the local number of particles in a quantum energy state $\mathcal{N}_{M}^{n}$ and the time $t$ as follow
\begin{equation}
    f=f(\mathcal{N}_M,\mathcal{N}_M^{n},t),
\end{equation}
where $\mathcal{N}_{M}^{n}$ is defined as
\begin{equation}
    \mathcal{N}_{M}^{n}= f_{nM} \delta \epsilon_n.
\end{equation}
Besides the local number of particles is a function of time $\mathcal{N}_J=\mathcal{N}_J(t)$ and the local number of particles in a quantum energy state is a function of time and the local number of particles $\mathcal{N}_J^{j}=\mathcal{N}_J^{j}(\mathcal{N}_M,t)$. This is due to when a cell transfer particles to another cell, the particles in some defined state energy levels must change.\\
In order to obtain the analogous transport equation to this system, we apply the variation method to the functional $\Ss$. Due to $\Ss$ in equilibrium is proportional to entropy, and therefore a maximum, the variation must be zero. Then the equation of motion is
\begin{equation}
    0=\sum_{M=1}^{K} \sum_n \ln \left[ \frac{f_{nM}}{1\mp f_{nM}} \right] \left[ \frac{\partial f_{nM}}{\partial t}+\frac{\partial f_{nM}}{\partial \mathcal{N}_J}\dot{\mathcal{N}}_J+\frac{\partial f_{nM}}{\partial \mathcal{N}_J^{j}}\left( \frac{\partial \mathcal{N}_J^{j}}{\partial t}+\frac{\partial \mathcal{N}_J^{j}}{\partial \mathcal{N}_L}\dot{\mathcal{N}}_L \right) \right]. \label{transportequation1}
\end{equation}
This equation is valid in equilibrium. When the system is out of equilibrium, the variation is different from zero. Due to this, an expression has to be included in (\ref{transportequation1}). This term, in analogy to the Boltzmann transport equation, will be the collision term and will be expressed as $\left( \frac{df_{nM}}{dt} \right)_{coll}$. Moreover, we propose that this term contributes to the equation of motion per each cell.\footnote{The collision term is only in a defined cell and local group of energy. If we are describing the total system, we need to include all contributions (each cell and energy group).} Therefore, (\ref{transportequation1}) casts into
\begin{eqnarray}
    \sum_{M=1}^{K}\sum_n \left( \frac{df_{nM}}{dt} \right)_{coll}&=&\sum_{M=1}^{K} \sum_n \ln \left[ \frac{f_{nM}}{1\mp f_{nM}} \right] \nonumber \\
    &&\times \left[ \frac{\partial f_{nM}}{\partial t}+\frac{\partial f_{nM}}{\partial \mathcal{N}_J}\dot{\mathcal{N}}_J+\frac{\partial f_{nM}}{\partial \mathcal{N}_J^{j}}\left( \frac{\partial \mathcal{N}_J^{j}}{\partial t}+\frac{\partial \mathcal{N}_J^{j}}{\partial \mathcal{N}_L}\dot{\mathcal{N}}_L \right) \right].  \nonumber \\ \label{transportequation2}
\end{eqnarray}
The equation (\ref{transportequation2}) is the time evolution equation to a quantum ideal gas. \\
\textcolor{red}{We obtained some results from the Statistical Mechanic in Quantum Mechanics scheme. However, the correspondence principle is important to recover all results from Classical Mechanics. This correspondence principle will be discussed in the next section.}

@@

\section{Correspondence principle}
%The variational entropy $H$ that was defined in (\ref{entropy2}) can be reduce to the Boltzmann's $H$ functional in the appropriated limit.\\
We begin defining the distribution function as the number of particles in a defined cell and an energy level $N_{nM}$ inside a volume $\de V_M \delta \epsilon_n$ as
\begin{equation}
    f_{nM}=\frac{N_{nM}}{ \de V_M \delta \epsilon_{n} }.
\end{equation}
Using this definition on the variational entropy (\ref{entropy2}) we obtain
\begin{eqnarray}
    H&=& \sum_{M=1}^{K} \sum_n
    \left[  
           \frac{N_{nM}}{ \de V_M\delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{nM}}{ \de V_M\delta \epsilon_{n}}
           \right)\pm 
           \left(  
                  1\mp \frac{N_{nM}}{ \de V_M \delta \epsilon_{n}}
           \right) \ln 
           \left(  
                   1\mp \frac{N_{nM}}{ \de V_M \delta \epsilon_{n}}
           \right)
    \right] \de V_M \delta \epsilon_{n}. \nonumber \\
    \label{h-quantic} 
\end{eqnarray}
In the classic limit, the particles occupy high energy levels, and the number of particles in most of the groups of state $n$ is small compared with the number of the elementary states $\epsilon_n$ of the group, that is
\begin{equation}
    \frac{N_{nM}}{ \de V_M \delta \epsilon_{n} } \approx 0,
\end{equation}
and with the previous approximation, the second term in (\ref{h-quantic}) is negligible, then (\ref{h-quantic}) yields
\begin{eqnarray}
    H&=& \sum_{M=1}^{K} \sum_n
    \left[  
           \frac{N_{nM}}{ \de V_M \delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{nM}}{ \de V_M \delta \epsilon_{n}}
           \right)
    \right] \de V_M \delta \epsilon_{n}. \label{h-quantic2}
\end{eqnarray}
In the case when the system is spatial-homogeneous, the definition of the distribution function is 
\begin{equation}
    f_{n}=\frac{N_{n}}{ \delta \epsilon_{n} },
\end{equation}
and using the same limit we obtain
\begin{eqnarray}
    H&=& \sum_{M=1}^{K} \sum_n
    \left[  
           \frac{N_{n}}{ \delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{n}}{ \delta \epsilon_{n}}
           \right)
    \right]  \de V_M \delta \epsilon_{n} = V \sum_n
    \left[  
           \frac{N_{n}}{ \delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{n}}{ \delta \epsilon_{n}}
           \right)
    \right] \delta \epsilon_{n} \nonumber \\
    &=& V \sum_n \left[N_n \ln N_n - N_n \ln \delta \epsilon_n  \right]=\de V_M K H_{Boltz},\label{h-quantic4}
\end{eqnarray}
where $B_{Boltz}$ now is (\ref{CH2}) expressed is the form of (\ref{reduce-h}) shown by Tolman.\footnote{See \cite{tolman} eq. (102.6) for more information.}\\
We can observe that if the number of particles is too big compared to the number of cells $K$, then we can conclude that the sum of the local variational entropy is equal to the variational entropy of the entire system.

@@

\section{Conclusions}
We conclude the following statements:\\
\begin{enumerate}
    \item Any system (classical or quantum) out of equilibrium (including inhomogeneities) evolves to the equilibrium state without using the transport equation as long as the local equilibrium hypothesis holds.
    \item We can obtain the transport equation from the $H$ functional.
    \item The variational entropy is proportional to the Boltzmann's $H$ functional in the limit of the high energy levels.
    \item We find an expression for the entropy to quantum gases.
    \end{enumerate}

%  \begin{equation}
%      \mathcal{H}
%  \end{equation}
